Estimating SARS-CoV-2 transmission in England from randomly sampled prevalence surveys

### 2024-03-31 incidence-prevalence.tex
proof

I start by defining these terms in \cref{inc-prev:sec:definitions}.
I then move on to explaining the three processes that link the incidence, prevalence, and the data: the \emph{infection process}, the \emph{prevalence process}, and the \emph{observation process}.

In (\cref{inc-prev:sec:infection-process}) I introduce the infection process, describing the infections' generating mechanism; in (\cref{inc-prev:sec:prevalence-process}) the prevalence process, which generates prevalence from the incidence; and in (\cref{inc-prev:sec:observation-process}), the observation process, which generates the data from the prevalence through the sampling mechanism of a population prevalence survey, guided by the design of the CIS.

@@

\emph{Prevalence} of a disease on day $t$, is the proportion of people with the disease on that day~\autocite[90]{lashModern}.
I refer to individuals with the disease as \emph{prevalent} individuals.
Whether an individual is prevalent is often expensive or error-prone to measure and hence is a latent quantity.
Furthermore, what is meant by ``having'' a disease can depend on context; for example, it could be the presence of symptoms, the ability to transmit, or detectable status (the difference between these was discussed in \cref{E-biology-data:sec:natural-history})
Within this thesis, I define an individual as prevalent on day $t$ if they are detectable (detectable is defined in \cref{E-biology-data:sec:detectable}).
I assume that an individual's prevalent status cannot change over the course of a day.
Therefore, prevalence is constant over any single day $t$.
I denote the number of prevalent individuals on day $t$ as $P_t$; hence, the prevalence is $P_t/\Npop$.

\emph{Duration}, or \emph{time-to-event}, distributions are expressed in several equivalent ways~\autocite[17]{sunStatisticala}.
The duration distribution I will be concerned with is the distribution of the duration of detectability.
Consider an individual $i$ infected on day $B_i$ that is prevalent during the interval $[B_i, E_i]$ (as in \cref{E-biology-data:sec:cis-episodes}).
I define this period as an \emph{infection episode}, beginning on day $B_i$ and ending on day $E_i$; therefore, with a duration $D_i = E_i - B_i + 1$.
I model the $D_i$s as a discrete iid (independent and identically distributed) random variables; extending the methodology to account for covariates is left to further work.
The duration is a positive integer, \ie $D_i \in \{1, 2, \dots, \dmax \}$, where $\dmax$ is the maximum duration.
The distribution of $D_i$ can be expressed through the standard probability mass function (pmf), $f_{D}(d) = \prob(D = d)$; or cumulative density function (CDF), $F_{D}(d) = \prob(D \leq d) = \sum_{i=1}^d f_{D}(i)$.
Alternatively, the \emph{survival function} gives the probability that the duration is at least $d$ days: $S(d) = \prob(D \geq d) = 1 - F_{D}(D - 1)$ with $S(1) = 1$.
%(a common alternative convention for survival analysis with discrete variables is to define the survival function as $\prob(D >d)$).
The \emph{hazard} on day $d$ is the probability of recovering on day $d$ conditional on still being positive on day $d$: $\lambda(d) = \prob(D = d \mid D \geq d) = f_{D}(d) / S(d)$.
Throughout this thesis, the survival function will be the primary quantity of interest, and normally parameterized in terms of the hazard; these two quantities are related by $S(d) = \prod_{i=1}^{d-1} (1 - \lambda(i))$.

Since $D_i$ is a discrete random variable, $0 \leq f_D(d) \leq 1$ for all $d$ and $\sum_{k=1}^{\dmax} f_D(k) = 1$.
By definition, $F_D(d) \in [0, 1]$ and increases monotonically.
Therefore, $S(d) \in [0, 1]$ and decreases monotonically.
Similarly, $\lambda(d) \in [0,1]$, but can otherwise vary arbitrarily (\ie there is no monotonic or sum constraint).
The lack of monotonicity or sum constraint on the hazard makes it an attractive parameterization for inference because these constraints are challenging to implement in many inference frameworks~\autocite{heBayesian}.

@@

First, in \cref{inc-prev:sec:poisson-process}, I introduce phenomenological models, where the infection process is a \emph{Poisson process}.
A Poisson process is a widely used type of counting process both within and outside of epidemiology.
It makes no explicit assumptions on the underlying biological or epidemiological mechanism and can be a flexible model.

Afterwards, in \cref{SEIR:sec:mechanistic-models}, I introduce \emph{mechanistic models}.
Mechanistic models explicitly describe the population and transmission between individuals.
In particular, they include the feedback loop inherent in infectious diseases, where increased infections at a time $t$ lead to more infections at times $> t$.
They vary greatly in their details and realism~\autocite{murilloMultiscale}.

@@

Individuals in the susceptible compartment are those that could be infected (they have no immunity).
The proportion of the population that is susceptible at time $t$ is $s(t)$.
Individuals in the infectious compartment can infect others, and eventually recover.
The proportion of the population that is infectious at time $t$ is $i(t)$.
Individuals in the recovered compartment are immune to the disease and cannot be infected again.
The proportion of the population that have recovered by time $t$ is $r(t)$.
The model's state is $\vec{x}(t) = (s(t), i(t), r(t))^T$.

The SIR model is described by the following system of ODEs (ordinary differential equations)~\autocite[19]{keelingModeling}:
\begin{align}
\frac{ds(t)}{dt} &= -\lambda(t) s(t) \\
\frac{di(t)}{dt} &= \lambda(t) s(t) - \gamma i(t) \\
\frac{dr(t)}{dt} &= \gamma i(t)
\end{align}
where $\lambda(t) = \beta i(t)$, is the \emph{force of infection}: the risk of infection for a susceptible individual at time $t$~\autocite[17]{keelingModeling}.
The dynamics of the model are described by the parameters $\vec{\theta} = (\beta, \gamma)^T$, where
$\beta$ is the transmission rate, absorbing several terms as described below;
$\gamma$ is the recovery rate, the rate that infectious individuals that recover.
Hence, $1/\gamma$ is the mean infectious period~\autocite[367]{keelingModeling}.
These parameters are assumed to be constant over time.

@@

A few properties and assumptions of the SIR model are worth noting.
\begin{itemize}
    \item For any short period of time when the change in $s(t)$ is negligible, the solution to the system of ODEs includes exponential growth in the number of infectious individuals: $i(t) = i_0 e^{\psi t}$ for some constants $i_0$ and $\psi$~\autocite[section 1.2]{diekmannMathematical}.
    A notable example is the early part of an epidemic when $s(t) \approx 1$.
    This is the phase of initial exponential growth of an epidemic, producing the characteristic early epidemic curves.
    It is in this phase that $\R$ provides a good approximation of the epidemic dynamics.
    \item $\R = \beta / \gamma$ in the SIR model~\autocite[20]{keelingModeling}.
    This is the mean number of infections caused by an infectious individual per day, $\beta$, multiplied by the mean number of days for which they are infectious, $1/\gamma$.
    \item The effective reproduction number for this model is $\Re = \beta / \gamma s(t) = \R s(t)$~\autocite{pellisEstimation}.
    $1-s(t)$ is the fraction of infections that are avoided due to the presence of immune individuals in the population.
    \item The model assumes a closed population.
    This assumption implies several attributes.
    First, the total population size is constant.
    Second, imported infections are negligible. 
    Third, births and deaths of individuals are negligible over the length of time considered.
    Model extensions allow these factors in contexts where they are relevant~\autocites[26]{keelingModeling}[214]{kretzschmarMathematical}.
    \item There is no analytical solution to the system of ODEs.
      Solving the system requires numerical methods~\autocite[25]{keelingModeling}, and uses the parameters, $\vec{\theta}$, and initial state of the system, $\vec{x}(0)$.
    \item The model assumes immunity lasts until the end of the period considered~\autocite[61]{andersonInfectious}.
    In many contexts, including this thesis, this assumption is justified because the period is shorter than the length of immunity conferred~\autocite{milneImmunity}.
    For longer time periods, the model can be expanded to include waning immunity~\autocite[40]{keelingModeling}.
    \item A discretised incidence proportion for day $t$ can be defined as $Z_t / \Npop = s(t) - s(t+1)$.
        However, this will not in general give an integer value to $Z_t$.
        In practice, this is not an issue since a continuous approximation for the incidence will suffice.
\end{itemize}

@@

The basic SIR model is the deterministic approximation of a stochastic model where the duration of time in the infectious compartment is exponentially distributed~\autocite[96]{keelingModeling}.
This is an unrealistic assumption for most diseases and would lead to underestimating of the reproduction number~\autocite{lloydRealistic,wearingAppropriate}.
A standard extension to the SIR model is to instead assume a gamma distribution~\autocite[94]{keelingModeling,andersonSpread}.
The extension adds multiple infectious compartments, shown in \cref{SEIR:fig:SIR-gamma}.
The ODEs are now:
\begin{align}
\frac{ds(t)}{dt} &= -\lambda(t) s(t)\\
\frac{di_1(t)}{dt} &= \lambda(t) s(t) - n\gamma i_1(t) \\
\frac{di_2(t)}{dt} &= n\gamma i_1(t) - n \gamma i_2(t) \\
&\vdots \nonumber \\
\frac{di_n(t)}{dt} &= n\gamma i_{n-1}(t) - n \gamma i_n(t) \\
\frac{dr(t)}{dt} &= n\gamma i_n(t) \\
\lambda(t) &= \beta \sum_{j=1}^n i_j(t)
\end{align}
where $n$ is the number of infectious compartments.
The additional compartments added do not, in general, correspond to any biological feature or have a biological interpretation; they are purely a mathematical device to modify the distribution of time between becoming and ceasing to be infectious~\autocite{lloydRealistic}.
In this model the infectious period is distributed $\GamDist(n, n\gamma)$, a realistic model for a variety of diseases~\autocite{wearingAppropriate}.
This parameterization is chosen to give $1/\gamma$ as the mean infectious period (further details on the Gamma distribution are in \cref{E-distributions}).
A gamma distribution with the first parameter equal to 1 is an exponential distribution, in which case this model collapses to the basic SIR model in \cref{SEIR:sec:SIR}.
As $n$ increases, the between-individual variability in infectious period length reduces, tending to a constant as $n \to \infty$~\autocite{lloydRealistic}.

@@

The reproduction number in terms of the parameters is unchanged by this extension, \ie $\R = \beta / \gamma$, however, the growth rate in the exponential phase will be slower than before~\autocite[41]{keelingModeling}.

@@

The state vector is now a vector of vectors, one for each of the compartments.
For example, in the basic SIR model, $\vec{x}(t) = (\vec{s}(t), \vec{i}(t), \vec{r}(t))^T$.
The $a$th element of each vector ($a \in \set{A}$) is the proportion of strata $a$ in that compartment.
For example, $\vec{s}(t) = (s_1(t), \dots, s_A(t))^T$ where $s_a(t)$ is the proportion of the $a$th strata that are susceptible at time $t$.
Hence, $N_a s_a(t)$ is the number of susceptible people in stratum $a$, where $N_a$ is the number of individuals in stratum $a$.
These vectors replace the scalars in the previous equations.
Other, equivalent formulations of this model exist within the literature.
For example, defining $s_a$ as the proportion of the total population that are both in strata $a$ and susceptible or parameterized using numbers of individuals~\autocite[57]{keelingModeling}, rather than as the proportion of strata $a$ as I have done here.

The interactions between strata are included by replacing the scalar $\beta$ with a WAIFW (\emph{Who Acquires Infection From Whom}) matrix, $\matr{\beta}$~\autocite[58]{keelingModeling}.
The force of infection is now a vector, defined as $\vec\lambda(t) = \matr{\beta} \sum_{j=1}^n \vec{i}_j(t)$ where $\vec{i}_j$ is a vector such that $i_{j,a}$ is the proportion of stratum $a$ in the $j$th infectious state.
Their interpretation can be arrived at by considering a single infectious individual in strata $a'$, analogous to the single stratum case.
Assume any such individual comes into contact with individuals in strata $a$ at a rate $\kappa_{aa'}$ per day.
% Further, assume that each contact is with an individual chosen uniformly at random from within the strata.
% For such a randomly selected individual, the probability that they are susceptible is $s_a(t)$.
% Each of the infectious individual's contacts is therefore with a susceptible individual with probability $s(t)$, by the well-mixed assumption.
Following the same argument as in \cref{SEIR:sec:SIR} finds that the infectious individual makes contact with susceptible individuals in stratum $a$ at a mean rate of $\kappa_{aa'} s_a(t)$ per day.
Now assume that there is a constant probability of transmission upon contact between an infectious individual in strata $a'$ and susceptible in strata $a$ individual, $q_{aa'}$.
Therefore, the mean rate of infections in strata $a$ per day caused by this infectious individual is $\kappa_{aa'} s_a(t) q_{aa'}$ per day.
Multiplying by the number of infectious individuals in strata $a'$, $N_{a'} i_{a'}(t)$, and dividing by the size of strata $a$, $N_a$, gives $\kappa_{aa'} q_{aa'} s_a(t) \frac{1}{N_a} N_{a'} i_{a'}(t)$ where $i_{a'}(t) = \sum_{j=1}^n i_{j,a'}$ is the total proportion of stratum $a'$ that is infectious, regardless of which compartment they are in.
The constant terms here are absorbed into the WAIFW matrix, hence $\beta_{aa'} = \kappa_{aa'} q_{aa'} \frac{1}{N_a} N_{a'}$~\autocite[section 9.2]{diekmannMathematical}.
This gives the total force of infection on strata $a$ from all strata is $\lambda_a(t) = \sum_{a' \in \set{A}} \beta_{aa'} i_{a'}(t)$ as above.
If all $\beta_{aa'}$ are equal, then the model is equivalent to the homogeneous model (\ie without any structure).
The rate at which individuals move between the susceptible and exposed compartments is now $\vec{s}(t) \circ \vec\lambda(t)$, where $\circ$ denotes element-wise multiplication.

@@

Reproduction numbers in models with structured populations are more complex than the homogeneous case.
The WAIFM allows calculation of the number of secondary infections generated by a single infection in the $a$th strata,  this follows directly from its definition as $\frac{1}{\gamma} \sum_{a' \in \set{A}} \beta_{aa'}$. 
However, how to average these contributions to calculate the basic reproduction number is not obvious.
The standard solution to this was proposed by \textcite{diekmannDefinition} (see \textcite[chapter 7]{diekmannMathematical} for a gentler, but still rigorous, explanation).
First, they show that the distribution of infections across the strata during the initial period of exponential growth (when $\vec{s} \approx \vec{1}$) relies only on $\matr{\beta}$, and not the starting conditions.
This distribution is proportional to the \emph{dominant eigenvector} (the eigenvector corresponding to the largest eigenvalue) of $\matr{\beta}$.
Denote by $D(\matr{\beta})$ the \emph{dominant eigenvalue} (the largest eigenvalue) of $\matr{\beta}$.
The basic reproduction number is then defined as $\R = D(\matr{\beta}) / \gamma$.
This $\R$ keeps the same interpretation as in the homogeneous case, \ie the average number of secondary infections caused by a single infection in a fully susceptible population, where the average is taken over the distribution implied by the dominant eigenvector.
The important threshold that exponential growth occurs if and only if $\R > 1$ remains.

@@

The relationship between incidence and prevalence can be derived by considering the probability that an individual is prevalent at any time following an infection.
%I define the period the individual is prevalent for as their \emph{infection episode}.
%I index infection episodes with $i$.
%Denote the time that the infection episode begins as $B_i$ and the time it ends as $E_i \geq B_i$.
%The individual is prevalent during the interval $[B_i, E_i]$ (\ie including $B_i$ and $E_i$).
%The duration of the infection episode, the number of days for which they are prevalent, is $D_i = E_i - B_i + 1$.
%I assume that the $D_i$s are iid, with a discrete distribution defined by $\prob(D_i = d) = f_D(d)$ for $d = 1, \dots, d_\text{max}$, for some maximum episode length $d_\text{max}$, and $\prob(D_i = d) = 0$ otherwise.
%Denote by $F_D$ the cdf of $D_i$, that is $F_D(d) = \prob(D_i \leq d)$.
From this, we can derive the relationship between incidence and prevalence.

@@

In general, all the random variables in this section are dependent.
This is because they are linked by the $Z_t$s, which are themselves dependent (as explained in \cref{inc-prev:sec:infection-process}).
However, if we condition on the vector of incidence, $\vec{Z} = (Z_1, Z_2, \dots)^T$, then we will induce some independence.
This reduces the stochasticity to only that attributable to the prevalence process.
In particular, because the duration of infection episodes is independent, $R_{t_z,t_p}$ and $R_{t_z',t_p'}$ will be independent if $t_z \neq t_z'$; importantly, this holds even if $t_p = t_p'$.
Therefore, $P_{t_z,t_p}$ and $P_{t_z',t_p'}$ will be independent if $t_z \neq t_z'$.

@@

The expectation in \cref{inc-prev:eq:EPt} is a discrete convolution equation in the same form as \cref{E-intro:eq:disease-model} with $\zeta = 1$.
% The equation is of the same form as the equation used for HIV/AIDS backcalculation, as well as being found in many other fields~\autocite[and references therein]{brookmeyerBackcalculation}.

>>>>

I start by defining these terms in \cref{inc-prev:sec:definitions}.
I then explain the three relevant models: one for the infection process (\cref{inc-prev:sec:infection-process}); one for the prevalence process, the relevant disease model (\cref{inc-prev:sec:prevalence-process}) linking infections to prevalence; and finally the observation process for a prevalence survey like the CIS (\cref{inc-prev:sec:observation-process}).

@@

\emph{Prevalence} of a disease on day $t$, is the proportion of people with the disease on that day~\autocite[90]{lashModern}.
I refer to individuals with the disease as \emph{prevalent} individuals.
What is meant by ``having'' a disease can depend on context; for example, it could be the presence of symptoms, the ability to transmit, or detectable status (the difference between these was discussed in \cref{E-biology-data:sec:natural-history})
Within this thesis, I define an individual as prevalent on day $t$ if they are detectable (detectable is defined in \cref{E-biology-data:sec:detectable}).
I assume that an individual's prevalent status cannot change over the course of a day.
Therefore, prevalence is constant over any single day $t$.
I denote the number of prevalent individuals on day $t$ as $P_t$; hence, the prevalence is $P_t/\Npop$.
$P_t$ corresponds with $W_t$ in \cref{E-intro:eq:disease-model}.

\emph{Duration}, or \emph{time-to-event}, distributions are expressed in several equivalent ways~\autocite[17]{sunStatisticala}.
The duration distribution I will be concerned with is the distribution of the duration of detectability.
Consider an individual $i$ infected on day $B_i$ that is prevalent during the interval $[B_i, E_i]$ (as in \cref{E-biology-data:sec:cis-episodes}).
I define this period as an \emph{infection episode}, beginning on day $B_i$ and ending on day $E_i$; therefore, with a duration $D_i = E_i - B_i + 1$.
I model the $D_i$s as a discrete iid (independent and identically distributed) random variables; extending the methodology to account for covariates is left to further work.
The duration is a positive integer, \ie $D_i \in \{1, 2, \dots, \dmax \}$, where $\dmax$ is the maximum duration.
The distribution of $D_i$ can be expressed through the standard probability mass function (pmf), $f_{D}(d) = \prob(D = d)$; or cumulative density function (CDF), $F_{D}(d) = \prob(D \leq d) = \sum_{i=1}^d f_{D}(i)$.
Alternatively, the \emph{survival function} gives the probability that the duration is at least $d$ days: $S(d) = \prob(D \geq d) = 1 - F_{D}(D - 1)$ with $S(1) = 1$.
%(a common alternative convention for survival analysis with discrete variables is to define the survival function as $\prob(D >d)$).
The \emph{hazard} on day $d$ is the probability of recovering on day $d$ conditional on still being positive on day $d$: $\lambda_{d} = \prob(D = d \mid D \geq d) = f_{D}(d) / S(d)$.
Throughout this thesis, the survival function will be the primary quantity of interest, and normally parameterized in terms of the hazard; these two quantities are related by $S(d) = \prod_{i=1}^{d-1} (1 - \lambda_{i})$.

Since $D_i$ is a discrete random variable, $0 \leq f_D(d) \leq 1$ for all $d$ and $\sum_{k=1}^{\dmax} f_D(k) = 1$.
By definition, $F_D(d) \in [0, 1]$ and increases monotonically.
Therefore, $S(d) \in [0, 1]$ and decreases monotonically.
Similarly, $\lambda_{d} \in [0,1]$, but can otherwise vary arbitrarily (\ie there is no monotonic or sum constraint).
The lack of monotonicity or sum constraint on the hazard makes it an attractive parameterization for inference because these constraints are challenging to implement in many inference frameworks~\autocite{heBayesian}.

@@

First, in \cref{inc-prev:sec:poisson-process}, I introduce phenomenological models, where the infection process is a \emph{Poisson process}.
A Poisson process is a widely used type of counting process both within and outside of epidemiology.

Afterwards, in \cref{SEIR:sec:mechanistic-models}, I introduce \emph{mechanistic models}.
Unlike phenomenological models, they include the feedback loop inherent in infectious diseases, where increased infections at a time $t$ lead to more infections at times $> t$.

@@

Individuals in the susceptible compartment are those that could be infected (they have no immunity).
The proportion of the population that is susceptible at time $t$ is $s(t)$.
Individuals in the infectious compartment can infect others, and eventually recover.
The proportion of the population that is infectious at time $t$ is $i(t)$.
Individuals in the recovered compartment are immune to the disease and cannot be infected again.
The proportion of the population that have recovered by time $t$ is $r(t)$.
The model's state is $\vec{x}(t) = [s(t), i(t), r(t)]^T$.

The SIR model is described by the following system of ODEs (ordinary differential equations)~\autocite[19]{keelingModeling}:
\begin{align}
\frac{ds(t)}{dt} &= -\lambda(t) s(t) \\
\frac{di(t)}{dt} &= \lambda(t) s(t) - \gamma i(t) \\
\frac{dr(t)}{dt} &= \gamma i(t) \\
\lambda(t) &= \beta i(t)
\end{align}
where $\lambda(t)$, is the \emph{force of infection}: the risk of infection for a susceptible individual at time $t$~\autocite[17]{keelingModeling}.
The dynamics of the model are described by the parameters $\vec{\theta} = [\beta, \gamma]^T$, where
$\beta$ is the transmission rate, absorbing several terms as described below;
$\gamma$ is the recovery rate, the rate that infectious individuals that recover.
Hence, $1/\gamma$ is the mean infectious period~\autocite[367]{keelingModeling}.
These parameters are assumed to be constant over time.

@@

A few properties and assumptions of the SIR model are worth noting.
\begin{itemize}
    \item For any short period of time when the change in $s(t)$ is negligible, the solution to the system of ODEs includes exponential growth in the number of infectious individuals: $i(t) = i_0 e^{\psi t}$ for some constants $i_0$ and $\psi$~\autocite[section 1.2]{diekmannMathematical}.
    A notable example is the early part of an epidemic when $s(t) \approx 1$.
    This is the phase of initial exponential growth of an epidemic, producing the characteristic early epidemic curves.
    It is in this phase that $\R$ provides a good approximation of the epidemic dynamics.
    \item $\R = \beta / \gamma$ in the SIR model~\autocite[20]{keelingModeling}.
    This is the mean number of infections caused by an infectious individual per day, $\beta$, multiplied by the mean number of days for which they are infectious, $1/\gamma$.
    \item The effective reproduction number for this model is $\Re = \beta / \gamma s(t) = \R s(t)$~\autocite{pellisEstimation}.
    $1-s(t)$ is the fraction of infections that are avoided due to the presence of immune individuals in the population.
    \item The model assumes a closed population.
    This assumption implies several attributes.
    First, the total population size is constant.
    Second, imported infections are negligible. 
    Third, births and deaths of individuals are negligible over the length of time considered.
    Model extensions allow these factors in contexts where they are relevant~\autocites[26]{keelingModeling}[214]{kretzschmarMathematical}.
    \item There is no analytical solution to the system of ODEs.
      Solving the system requires numerical methods~\autocite[25]{keelingModeling}, and uses the parameters, $\vec{\theta}$, and initial state of the system, $\vec{x}(0)$.
    \item The model assumes immunity lasts until the end of the period considered~\autocite[61]{andersonInfectious}.
    In many contexts, including this thesis, this assumption is justified because the period is shorter than the length of immunity conferred~\autocite{milneImmunity}.
    For longer time periods, the model can be expanded to include waning immunity~\autocite[40]{keelingModeling}.
    \item A discretised incidence proportion for day $t$ ($t$ an integer) can be defined as $Z_t / \Npop = s(t) - s(t+1)$.
        However, this will not in general give an integer value to $Z_t$.
        In practice, this is not an issue since a continuous approximation for the incidence will suffice.
\end{itemize}

@@

The basic SIR model is the mean-field approximation of a stochastic model where the duration of time in the infectious compartment is exponentially distributed~\autocite[96]{keelingModeling}.
This is an unrealistic assumption for most diseases and would lead to underestimating of the reproduction number~\autocite{lloydRealistic,wearingAppropriate}.
A standard extension to the SIR model is to instead assume a gamma distribution~\autocite[94]{keelingModeling,andersonSpread}.
The extension adds multiple infectious compartments, shown in \cref{SEIR:fig:SIR-gamma}.
The ODEs are now:
\begin{align}
\frac{ds(t)}{dt} &= -\lambda(t) s(t)\\
\frac{di_1(t)}{dt} &= \lambda(t) s(t) - n\gamma i_1(t) \\
\frac{di_2(t)}{dt} &= n\gamma i_1(t) - n \gamma i_2(t) \\
&\vdots \nonumber \\
\frac{di_n(t)}{dt} &= n\gamma i_{n-1}(t) - n \gamma i_n(t) \\
\frac{dr(t)}{dt} &= n\gamma i_n(t) \\
\lambda(t) &= \beta \sum_{j=1}^n i_j(t)
\end{align}
where $n$ is the number of infectious compartments.
The additional compartments added do not, in general, correspond to any biological feature or have a biological interpretation; they are purely a mathematical device to modify the distribution of time between becoming and ceasing to be infectious~\autocite{lloydRealistic}.
In this model the infectious period is distributed $\GamDist(n, n\gamma)$, a realistic model for a variety of diseases~\autocite{wearingAppropriate}.
This parameterization is chosen to give $1/\gamma$ as the mean infectious period (further details on the Gamma distribution are in \cref{E-distributions}).
A gamma distribution with the first parameter equal to 1 is an exponential distribution, in which case this model collapses to the basic SIR model in \cref{SEIR:sec:SIR}.
As $n$ increases, the between-individual variability in infectious period length reduces, tending to a constant as $n \to \infty$~\autocite{lloydRealistic}.

@@

The reproduction number in terms of the parameters is unchanged by this extension, \ie $\R = \beta / \gamma$, however, the growth rate in the initial exponential phase is slower than before~\autocite[41]{keelingModeling}.

@@

The state vector is now a vector of vectors, one for each of the compartments.
For example, in the basic SIR model, $\vec{x}(t) = [\vec{s}(t), \vec{i}(t), \vec{r}(t)]^T$.
The $a$th element of each vector ($a \in \set{A}$) is the proportion of strata $a$ in that compartment.
For example, $\vec{s}(t) = [s_1(t), \dots, s_A(t)]^T$ where $s_a(t)$ is the proportion of the $a$th strata that are susceptible at time $t$.
Hence, $N_a s_a(t)$ is the number of susceptible people in stratum $a$, where $N_a$ is the number of individuals in stratum $a$.
These vectors replace the scalars in the previous equations.
Other, equivalent formulations of this model exist within the literature.
For example, defining $s_a$ as the proportion of the total population that are both in strata $a$ and susceptible or parameterized using numbers of individuals~\autocite[57]{keelingModeling}, rather than as the proportion of strata $a$ as I have done here.

The interactions between strata are included by replacing the scalar $\beta$ with a WAIFW (\emph{Who Acquires Infection From Whom}) matrix, $\matr{\beta}$~\autocite[58]{keelingModeling}.
The force of infection is now a vector, defined as $\vec\lambda(t) = \matr{\beta} \sum_{j=1}^n \vec{i}_j(t)$ where $\vec{i}_j$ is a vector such that $i_{j,a}$ is the proportion of stratum $a$ in the $j$th infectious state.

$\vec{\lambda}_t$ can be arrived at by considering a single infectious individual in strata $a'$, analogous to the single stratum case.
Assume any such individual comes into contact with individuals in strata $a$ at a rate $\kappa_{aa'}$ per day.
% Further, assume that each contact is with an individual chosen uniformly at random from within the strata.
% For such a randomly selected individual, the probability that they are susceptible is $s_a(t)$.
% Each of the infectious individual's contacts is therefore with a susceptible individual with probability $s(t)$, by the well-mixed assumption.
Following the same argument as in \cref{SEIR:sec:SIR} finds that the infectious individual makes contact with susceptible individuals in stratum $a$ at a mean rate of $\kappa_{aa'} s_a(t)$ per day.
Now assume that there is a constant probability of transmission upon contact between an infectious individual in strata $a'$ and susceptible in strata $a$ individual, $q_{aa'}$.
Therefore, the mean rate of infections in strata $a$ per day caused by this infectious individual is $\kappa_{aa'} s_a(t) q_{aa'}$ per day.
Multiplying by the number of infectious individuals in strata $a'$, $N_{a'} i_{a'}(t)$, and dividing by the size of strata $a$, $N_a$, gives $\kappa_{aa'} q_{aa'} s_a(t) \frac{1}{N_a} N_{a'} i_{a'}(t)$ where $i_{a'}(t) = \sum_{j=1}^n i_{j,a'}$ is the total proportion of stratum $a'$ that is infectious, regardless of which compartment they are in.
The constant terms here are absorbed into the WAIFW matrix, hence $\beta_{aa'} = \kappa_{aa'} q_{aa'} \frac{1}{N_a} N_{a'}$~\autocite[section 9.2]{diekmannMathematical}.
This gives the total force of infection on strata $a$ from all strata is $\lambda_a(t) = \sum_{a' \in \set{A}} \beta_{aa'} i_{a'}(t)$ as above.
If all $\beta_{aa'}$ are equal, then the model is equivalent to the homogeneous model (\ie without any structure).
The rate at which individuals move between the susceptible and exposed compartments is now $\vec{s}(t) \circ \vec\lambda(t)$, where $\circ$ denotes element-wise multiplication.

@@

Reproduction numbers in models with structured populations are more complex than the homogeneous case.
The WAIFM allows calculation of the number of secondary infections generated by a single infection in the $a$th strata,  this follows directly from its definition as $\frac{1}{\gamma} \sum_{a' \in \set{A}} \beta_{a'a}$. 
However, how to average these contributions to calculate the basic reproduction number is not obvious.

The standard solution to this was proposed by \textcite{diekmannDefinition} (see \textcite[chapter 7]{diekmannMathematical} for a gentler, but still rigorous, explanation).
First, they show that the distribution of infections across the strata during the initial period of exponential growth (when $\vec{s} \approx \vec{1}$) relies only on $\matr{\beta}$, and not the starting conditions.
This distribution is proportional to the \emph{dominant eigenvector} (the eigenvector corresponding to the largest eigenvalue) of $\matr{\beta}$.
Denote by $D(\matr{\beta})$ the \emph{dominant eigenvalue} (the largest eigenvalue) of $\matr{\beta}$.
The basic reproduction number is then defined as $\R = D(\matr{\beta}) / \gamma$.
This $\R$ keeps the same interpretation as in the homogeneous case, \ie the average number of secondary infections caused by a single infection in a fully susceptible population, where the average is taken over the distribution implied by the dominant eigenvector.
The important threshold that, in the mean-field approximation, exponential growth occurs if and only if $\R > 1$ remains.

@@

The prevalence process is the disease model (see \cref{E-intro:sec:disease-model}) relating incidence and prevalence.
The relationship between incidence and prevalence can be derived by considering the probability that an individual is prevalent at any time following an infection.
%I define the period the individual is prevalent for as their \emph{infection episode}.
%I index infection episodes with $i$.
%Denote the time that the infection episode begins as $B_i$ and the time it ends as $E_i \geq B_i$.
%The individual is prevalent during the interval $[B_i, E_i]$ (\ie including $B_i$ and $E_i$).
%The duration of the infection episode, the number of days for which they are prevalent, is $D_i = E_i - B_i + 1$.
%I assume that the $D_i$s are iid, with a discrete distribution defined by $\prob(D_i = d) = f_D(d)$ for $d = 1, \dots, d_\text{max}$, for some maximum episode length $d_\text{max}$, and $\prob(D_i = d) = 0$ otherwise.
%Denote by $F_D$ the cdf of $D_i$, that is $F_D(d) = \prob(D_i \leq d)$.
In what follows, I use this to derive the relationship between incidence and prevalence.

@@

In general, all the random variables in this section are dependent.
This is because they are linked by the $Z_t$s, which are themselves dependent (as explained in \cref{inc-prev:sec:infection-process}).
However, if we condition on the vector of incidence, $\vec{Z} = [Z_1, Z_2, \dots]^T$, then we will induce some independence.
This reduces the stochasticity to only that attributable to the prevalence process.
In particular, because the duration of infection episodes is independent, $R_{t_z,t_p}$ and $R_{t_z',t_p'}$ will be independent if $t_z \neq t_z'$; importantly, this holds even if $t_p = t_p'$.
Therefore, $P_{t_z,t_p}$ and $P_{t_z',t_p'}$ will be independent if $t_z \neq t_z'$.

@@

The expectation in \cref{inc-prev:eq:EPt} is a discrete convolution equation in the same form as \cref{E-intro:eq:disease-model} with $\zeta = 1$ and $f(i) = S(i+1)$.
% The equation is of the same form as the equation used for HIV/AIDS backcalculation, as well as being found in many other fields~\autocite[and references therein]{brookmeyerBackcalculation}.

Considering multiple values of $t$ makes \cref{inc-prev:eq:EPt} into a system of simultaneous equations.
Therefore, the system can be written in matrix form:
\begin{align}
  \E(\vec{P} \mid \vec{Z}) = \matr{S} \vec{Z}. \label{inc-prev:eq:EPt-matrix} \\
\intertext{$\matr{S}$ is a lower-triangular matrix formed from $S$:}
    (\matr{S})_{i,j} &= \begin{cases}
        S(i - j + 1) & i - \dmax - 1 \leq j \leq i \\
        0 & \text{otherwise.}
    \end{cases} \label{inc-prev:eq:Smatrix}
\end{align}
where $\vec{P} = [P_1, P_2, \dots]$.
In \cref{E-backcalc}, I will need to solve for this system for $\vec{Z}$ given a known $\hat{P} = \E(\vec{P} \mid {Z})$.
Assuming $\vec{P}$ and $\vec{Z}$ have the same dimension then $\matr{S}$ is lower-triangular with strictly positive diagonal elements; therefore, $\matr{S}$ is invertible, and the system has a unique solution~\autocite[424]{bernsteinFacts}.
The lower-triangular structure of $\matr{S}$ makes allows solving the system through forward-substitution, an efficient method that avoids explicit matrix inversion~\autocite{cormenMatrix}.
Forward-substitution sequentially solves the $t$th element of $\vec{Z}$ ($t = 1, \dots, T$) by calculating:
\begin{align}
Z_t
&= \frac{\hat{P}_t - \sum_{i=1}^{t-1} S_{ti} Z_i}{S_{tt}} \\
&= \hat{P}_t - \sum_{i=1}^{t-1} S(t - i + 1) Z_i.
\label{inc-prev:eq:forward-substitute}
\end{align}
where $S(t)$ for $t > \dmax$ is defined to be 0.
